---
title:  "Introduction to State Space Time Series Analysis"
output: html_document
author: "Elias Ardila"
---

# Introduction

The standard regression model for $n$ observations of $y$ and $x$ is written as
$$
y_{i} = a + bx_{i} + \epsilon_{i},\quad \epsilon_{i} \sim \mathcal{N}(0,\sigma_{\epsilon}^{2})
$$

```{r loaddata}
ukksi <- scan("ukksi.txt")
logukksi <- log(ukksi)
x <- 1:192
```

```{r linearmodel}
mod1 <- lm(logukksi ~ x)
summary(mod1)
anova(mod1)
aov(logukksi ~ x)
```

A scatter plot of variable $y$ on $x$ together with the best fitting line
according to classical linear regression.

```{r plot1, fig.width=8, fig.height=7, fig.align='center',  out.width='.9\\linewidth'}
plot(logukksi,pch=3,xaxt="n",xlab="",ylab="",yaxt="n",cex=.3)
axis(1, at=seq(0,200,by=5),labels=rep("",length(seq(0,200,by=5))))
axis(1,at=seq(0,200,by=20),labels=seq(0,200,by=20))
axis(2,at=seq(7.0,7.9,by=0.1),labels=seq(7.0,7.9,by=0.1),las=1)
lines(fitted(mod1) ~ x,lty=3)
legend("topright",
       legend=c("log UK drivers KSI against time (in months)","regression line"),
       lty=c(NA,3),pch=c(3,NA))
```

The equation of the regression line is
$$
\hat{y_i} = \hat{a} + \hat{b}x_i = `r round(coef(mod1)[1],4) `  + (`r round(coef(mod1)[2],5) ` x_i )\\
$$
with error variance $\sigma_{\epsilon}^{2} = `r summary(mod1)$sigma^2 `$.

```{r sigma}
n <- length(ukksi)
n
S <- t(logukksi - coef(mod1)[1] - coef(mod1)[2]*x)%*%(logukksi - 
                                  coef(mod1)[1] - coef(mod1)[2]*x)
s2hat <- as.numeric(S/(n-2))
s2hat
```

The standardt F-test yields $F=`r round(summary(mod1)$fstatistic[1],3) `$.


# Predicted Values and Residuals
```{r resid}
ei <- residuals(mod1)
```

Log of the number of UK drivers KSI plotted as time series.

```{r plot2, fig.width=8, fig.height=7, fig.align='center',out.width='.9\\linewidth'}
plot(logukksi,pch=3,xaxt="n",xlab="",ylab="",yaxt="n",type="l")
axis(1,at=seq(0,200,by=20),labels=seq(0,200,by=20))
axis(2,at=seq(7.0,7.9,by=0.1),labels=seq(7.0,7.9,by=0.1),las=1)
legend("topright",
       legend=c("log UK drivers KSI"),
       lty=c(1))
```

Residuals of classical linear regression of the log of the number of
UK drivers KSI on time.

```{r plot3, fig.width=8, fig.height=7, fig.align='center',out.width='.9\\linewidth'}
plot(ei,pch=3,xaxt="n",xlab="",ylab="",yaxt="n",type="l",lty=3)
axis(1,at=seq(0,200,by=20),labels=seq(0,200,by=20))
axis(2,at=round(seq(-0.3,0.4,by=0.1),1),labels=round(seq(-0.3,0.4,by=0.1),1),las=1)
legend("topleft",
       legend=c("residuals"),
       lty=3)
abline(h=0)
```

Correlogram of classical regression residuals.

```{r plot5, fig.width=8, fig.height=7, fig.align='center',out.width='.9\\linewidth'}
acf(ei,lag.max=15,main="ACF-regression residuals")
##abline(h=c(-ll,ll),lty=3)
```

```{r manual}
yt <- logukksi
as.numeric(t(yt)%*%yt)
yt.hat <- fitted(mod1)

as.numeric(t(yt.hat + ei)%*%(yt.hat + ei))
ss.model <- as.numeric(t(yt.hat)%*%yt.hat)
ss.model
ss.res <- as.numeric(t(ei)%*%ei)
ss.res
ss.model + ss.res
ss.total.unc <- as.numeric(t(yt)%*%yt)
ss.total.unc

b0 <- coef(mod1)[1]
b1 <- coef(mod1)[2]

## ss.model
n*mean(yt)^2 + (b1^2)*as.numeric(t(x - mean(x))%*%(x - mean(x)))

ss.total.unc - ss.model

ss.res

ss.regr <- (b1^2)*as.numeric(t(x - mean(x))%*%(x - mean(x)))
ss.regr

ss.regr + ss.res
```
